Automated Formal Synthesis\\ of Safe Digital Controllers for Continuous Plants

### 2019-01-13 automatica/paper.tex
changes

The technique underpinning tools such as Pessoa~\cite{mazo2010pessoa}, related extensions \cite{Anta2010,liu16} and applications \cite{zamani2014}, 
synthesizes correct-by-design embedded control software for LTI models.    
It is based on the abstraction of the model to a finite-state machine, 
and on the formal control synthesis over safety- and reachability-based temporal properties thereon.  
%Based on this safety specification, \mbox{Pessoa} can synthesize embedded controller software for a range of
%properties.  
The controller software can then be refined (implemented) over the concrete LTI model. 
%can be more complicated than
%the state-feedback control we synthesize, and the properties available cover more detail.  
However, relying on state-space discretization \mbox{Pessoa} is likely to incur in scalability limitations.   
%Along this research line, \cite{Anta2010,liu16} studies the synthesis of digital controllers for
%continuous dynamics, and \cite{zamani2014} extends the approach to the Network Control Systems setup.

@@

There are two different problems that arise from FWL semantics.  The first
is the error caused by the inability to represent the exact
state of the physical system, while the second relates to rounding and
saturation errors during mathematical operations.  
In~\cite{fialho1994stability}, a stability measure based on the error of the digital dynamics ensures that
the deviation introduced by FWL does not lead to instability.  
\cite{DBLP:journals/automatica/WuLCC09} uses $\mu$-calculus to directly model the digital controller so that selected
parameters result by design in stable dynamics.  
The analyses in~\cite{DBLP:conf/hybrid/RouxJG15, DBLP:conf/hybrid/WangGRJF16} rely on an
invariant computation on the discrete dynamics using Semi-Definite
Programming (SDP):  while the former contribution uses bounded-input and bounded-output
(BIBO) properties to determine stability, the latter uses Lyapunov-based
quadratic invariants.  In both cases, the SDP solver uses floating-point
arithmetic and soundness is checked by bounding the error.  
An alternative is~\cite{park2016scalable}, where the verification of given control code is
performed against a known model by extracting an LTI model of the code by
symbolic execution: to account for rounding errors, an upper bound is
introduced in the verification phase.  
The work in~\cite{picasso2003stabilization} introduces invariant sets as a mechanism to
bound the quantization error effect on stabilization.  
% as an invariant set that always converges toward the controllable set.  
Similarly, 
\cite{liberzon2003hybrid} evaluates the quantization error dynamics and \note{[unclear] bounds its trajectory to a known region over a finite time period}.   
%This technique works for both linear and non-linear systems.

@@

The gain matrix $K$ can be set so that the closed-loop discrete dynamics are
shaped as desired, for instance according to a specific stability goal or
around a specific dynamical behavior \cite{astrom1997computer}.  As argued
later in this work, we will target more complex objectives, such as
quantitative safety requirements, which are not typical in the digital
control literature.  Further, we will embrace the digital nature of the
controller, which manipulates quantized signals as discrete quantities represented with finite precision.

@@

In this work we employ asymptotic stability in the CEGIS loop,  
as an objective for guessing controllers that are later proven sound over safety requirements.  
Asymptotic stability is a property that amounts to convergence of the model executions to an equilibrium point, 
starting from any states in a neighborhood of the point (see Figure~\ref{fig:ct} for the portrait of a stable execution, converging to the origin).   
In the case of linear systems as in~\eqref{eq:closedloopss}, 
considered with a zero reference signal, 
the equilibrium point of interest is the origin.

A discrete-time LTI system as \eqref{eq:closedloopss} is asymptotically
stable if all the roots of its characteristic polynomial (i.e., the
eigenvalues of the closed-loop matrix $A_d - B_d K$) are inside the unity
circle of the complex plane, i.e., their absolute values are strictly less than
one~\cite{astrom1997computer} (this simple sufficient condition can be generalised, 
however this is not necessary in our work).  
In this paper, we express this stability specification $\phi_\mathit{stability}$ in terms of a check known as
\emph{Jury's criterion}~\cite{fadali}: this is an easy algebraic formula to
select the entries of matrix $K$ so that the closed-loop dynamics are shaped
as desired.

@@

We are not limited to the synthesis of digital stabilizing controllers -- a
well known task in the literature on digital control systems -- but target
safety requirements with an overall approach that is sound and automated. 
More specifically, we require that the closed-loop system
\eqref{eq:closedloopss} meets given safety specifications.  A safety
specification gives raise to a requirement on the states of the model, such
that the feedback controller (namely the choice of the gains matrix~$K$)
must ensure that the state never violates the requirement.  Note that a
stable, closed-loop system is not necessarily a safe system: indeed, the
state values may leave the safe part of the state space while they converge
to the equilibrium, which is typical in the case of oscillatory dynamics. 
%However, a safe system will always be stable.  
In~this work, the safety property is expressed as:
%
\begin{equation}
\label{eq:safetyliteral}
\phi_\mathit{safety}\iff \forall k\ge 0.\, \bigwedge_{i=1}^{n}{\underline{x_{i}} \leq x_{i,k} \leq \overline{x_{i}}},
\end{equation}
%
%\addtodo{We consider $x_{i}^{-}$ and $x_{i}^{+}$ to be -1 and 1, respectively.}
%
where $\underline{x_{i}}$ and $\overline{x_{i}}$ are lower and upper bounds
for the $i$-th coordinate $x_{i}$ of state $x\in \mathbb R^n$ at the $k$-th
instant, respectively.  This means that the states will always be within an $n$-dimensional hyper-box.

@@

The key contribution of our work is that, when synthesising controllers, we precisely model the finite-precision arithmetic in the digital controller, guaranteeing that the controllers are safe. The finite precision arithmetic introduces the following errors, shown in Figure~\ref{fig:digitalsystem}. These errors are present in the real world, and we must model them accurately 
in order to capture the full behaviour of the system. Details of how we model this arithmetic are in Section~\ref{sec:}:

@@

\item The DAC could potentially introduce further errors. However, we assume that the DAC converts numbers to analogue signals with the same precision as the controller. 
As a result we can ignore any additional errors introduced by the DAC. Again, this assumption is reasonable based on commonly available hardware. 
\end{itemize}

@@

We consider an input specification of the form 
$$
\exists P .\, \forall a.\, \phi(a, P),
$$  
where $P$ ranges over functions (where a function is represented by the program computing it),
$a$ ranges over ground terms, 
and $\phi$ is a quantifier-free logical formula. 
We~interpret the ground terms over some domain~$\mathcal{D}$. 
This is a synthesis problem, where the objective is indeed to select a valid $P$ satisfying $\phi$ over all the $a$ terms.

@@

In the context of the formal synthesis of safe controllers, 
the set of possible {\sc inputs} corresponds to the set of possible initial states and the 
candidate program $P$ is a candidate controllers $K$. The synthesis block generates a candidate
controller that works for a subset of the possible initial states, and the verifier checks whether 
the controller works for all possible initial states.

@@

%-------------------------------
\subsection{Formal specification of stability} 
\label{ssec:stabspecification}
%-------------------------------

There are a number of procedures that can be used for stability analysis of dynamical models~\cite{daes20161, Bessa16}, 
which can fit our automatic verification engine.  
%one based on Schur's decomposition
%and another one based on Jury's criterion~\cite{astrom1997computer}.  
Here we choose Jury stability criterion~\cite{astrom1997computer}, 
in view of its efficiency and ease of integration within our implementation.  
This method checks the stability working in the complex domain of the characteristic polynomial $S(z)$,  
%
% of the matrix $\left( \begin{array}{c} Y \\ U \end{array}\right)$, where $Y(z)$ is the closed-loop system output and U(z) is the controller output. 
%% \red{[unclear: what matrix? We need to clarify the relationship
%% between the char. poly. $S(z)$ and the later quantities $\Delta
%% N_{G}(z), \Delta D_{G}(z)$ and corresponding controller's
%% structure.]}.
%
considered in its general form as 
% for $S(z)$:
%
\begin{equation*}
S(z) = a_0z^N+a_1z^{N-1}+\cdots+a_{N-1}z+a_N,\,\, a_0\neq0. 
\end{equation*}
An LTI system is stable if all the roots of its characteristic polynomial $S(z)$ (i.e., the 
eigen values of the closed loop matrix $A_d-B_dK$) are inside the unit circle of the complex plane, 
i.e., their absolute values are less than one.

@@

%-------------------------------
\subsection{Formal specification of safety} 
\label{ssec:safespecification}
%-------------------------------

@@

Furthermore, it is practically relevant to consider constraints $\phi_\mathit{input}$ on the input
signal $u_{k}$ and requirements $\phi_\mathit{init}$ on the initial states $x_0$, 
both of which we be expressed via given bounds as: 
$\phi_\mathit{input} := {\forall k\ge 0, \underline{u} \leq u_{k} \leq \overline{u}} $, 
and $\phi_\mathit{init} := \bigwedge_{i=1}^{n} \underline{x_{i,0}} \leq x_{i,0} \leq \overline{x_{i,0}}$. 
For the input signal, these requirements practically mean that the control input saturates in view of physical constraints.

@@

\fi
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Numerical representation and soundness} 
\label{sec:numeric_rep}
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

@@

Let us first consider the effect of the quantization errors on safety. 
Within the controller, state values are manipulated at low precision,
alongside the vector multiplication $Kx$.
The inputs are computed using the following equation: 
%
\begin{align*}
u_{k}&=-(\mathcal{F}_{\langle I_c,F_c \rangle}(K)\cdot\mathcal{F}_{\langle I_c,F_c \rangle}(x_{k})). 
\end{align*}

This induces two types of the errors detailed above: first, the truncation
error due to representing $x_k$ as $\mathcal{F}_{\langle I_c,F_c
\rangle}(x_{k})$; and second, the rounding error introduced by the
multiplication operation.

Previous studies~\cite{gangli1} show that the FWL affects the poles and
zeros positions, degrading the closed-loop dynamics, causing steady-state
errors and eventually de-stabilizing the system~\cite{Bessa16}.  However,
since in this paper we require stability only as a precursor to safety, it
is sufficient to check that the (perturbed, noisy) model converges to a
neighborhood of the equilibrium within the safe set.

In the following, we shall disregard these steady-state errors (caused by
FWL effects) when stability is ensured by synthesis, and then verify its
safety accounting for the finite-precision errors.

>>>>

The technique underpinning tools such as Pessoa~\cite{mazo2010pessoa}, related extensions \cite{Anta2010,liu16}, and applications \cite{zamani2014}
synthesizes correct-by-design embedded control software for LTI models.    
It is based on the abstraction of the model to a finite-state machine, 
and on the formal control synthesis over safety- and reachability-based temporal properties thereon.  
%Based on this safety specification, \mbox{Pessoa} can synthesize embedded controller software for a range of
%properties.  
The controller software can then be implemented over the concrete LTI model. 
%can be more complicated than
%the state-feedback control we synthesize, and the properties available cover more detail.  
However, relying on state-space discretization, \mbox{Pessoa} is likely to incur scalability limitations.   
%Along this research line, \cite{Anta2010,liu16} studies the synthesis of digital controllers for
%continuous dynamics, and \cite{zamani2014} extends the approach to the Network Control Systems setup.

@@

There are two different problems that arise from FWL semantics in digital controllers.  The first
is the error caused by the inability to represent the exact
state of the physical system, while the second relates to rounding and
saturation errors during mathematical operations.  
In~\cite{fialho1994stability}, a stability measure based on the error of the digital dynamics ensures that
the deviation introduced by FWL does not lead to instability.  
\cite{DBLP:journals/automatica/WuLCC09} uses $\mu$-calculus to directly model the digital controller so that selected
parameters result by design in stable dynamics.  
The analyses in~\cite{DBLP:conf/hybrid/RouxJG15, DBLP:conf/hybrid/WangGRJF16} rely on an
invariant computation on the discrete dynamics using Semi-Definite
Programming (SDP):  while the former contribution uses bounded-input and bounded-output
(BIBO) properties to determine stability, the latter uses Lyapunov-based
quadratic invariants.  In both cases, the SDP solver uses floating-point
arithmetic and soundness is checked by bounding the error.  
An alternative is~\cite{park2016scalable}, where the verification of given control code is
performed against a known model by extracting an LTI model of the code by
symbolic execution: to account for rounding errors, an upper bound is
introduced in the verification phase.  
The work in~\cite{picasso2003stabilization} introduces invariant sets as a mechanism to
bound the quantization error effect on stabilization.  
% as an invariant set that always converges toward the controllable set.  
Similarly, 
\cite{liberzon2003hybrid} evaluates the quantization error dynamics and calculates an upper and lower
bound for the possible trajectory of the system up to a finite time. 
%This technique works for both linear and non-linear systems.

@@

The gain matrix $K$ can be set so that the closed-loop discrete dynamics are
shaped as desired, for instance according to a specific stability goal or
around a specific dynamical behavior \cite{astrom1997computer}.  As argued
later in this work, we will target a more complex objective,
quantitative safety requirements, which are not typical in the digital
control literature. We will precisely account for the digital nature
of the controller, which manipulates quantised signals as discrete quantities represented with
finite precision.

% Further, we will embrace the digital nature of the
%controller, which manipulates quantized signals as discrete quantities represented with finite precision.

@@

In this work we employ asymptotic stability in the CEGIS loop,  
as a means for reducing the search space of possible safe controller. As discussed later, a safe controller is necessarily asymptotically stable, although the reverse is not true.
%as an objective for guessing controllers that are later proven sound over safety requirements.  
Asymptotic stability is a property that amounts to convergence of the model executions to an equilibrium point, 
starting from any states in a neighborhood of the point (see Figure~\ref{fig:ct} for the portrait of a stable execution, converging to the origin).   
In the case of linear systems %as in~\eqref{eq:closedloopss}, 
considered with a zero reference signal, 
the equilibrium point of interest is the origin.

A discrete-time LTI system %as \eqref{eq:closedloopss} 
is asymptotically
stable if all the roots of its characteristic polynomial (i.e., the
eigenvalues of the closed-loop matrix $A_d - B_d K$) are inside the unity
circle of the complex plane, i.e., their absolute values are strictly less than
one~\cite{astrom1997computer} (this simple sufficient condition can be generalised, 
however this is not necessary in our work).  
In this paper, we express this stability specification $\phi_\mathit{stability}$ in terms of a check known as
\emph{Jury's criterion}~\cite{fadali}: this is an easy algebraic formula to
select the entries of matrix $K$ so that the closed-loop dynamics are shaped
as desired.

@@

We are not limited to the synthesis of digital stabilizing controllers -- a
well known task in the literature on digital control systems -- but target
safety requirements with an overall approach that is sound and automated. 
More specifically, we require that the closed-loop system
\eqref{eq:closedloopss} meets given safety specifications.  A safety
specification gives rise to a requirement on the states of the model, such
that the feedback controller (namely the choice of the gain matrix~$K$)
must ensure that the state never violates the requirement.  Note that a
stable, closed-loop system is not necessarily a safe system: indeed, the
state values may leave the safe part of the state space while they converge
to the equilibrium, which is typical in the case of oscillatory dynamics. 
%However, a safe system will always be stable.  
In~this work, the safety property is expressed as:
%
\begin{equation}
\label{eq:safetyliteral}
\phi_\mathit{safety}\iff \forall k\ge 0.\, \bigwedge_{i=1}^{n}{\underline{x_{i}} \leq x_{i,k} \leq \overline{x_{i}}},
\end{equation}
%
%\addtodo{We consider $x_{i}^{-}$ and $x_{i}^{+}$ to be -1 and 1, respectively.}
%
where $\underline{x_{i}}$ and $\overline{x_{i}}$ are lower and upper bounds
for the $i$-th coordinate $x_{i}$ of state $x\in \mathbb R^n$ at the $k$-th
instant, respectively.  This requires that the states will always be within an $n$-dimensional hyper-box.

@@

The key contribution of our work is that, when synthesising controllers, we precisely model the finite-precision arithmetic in the digital controller, guaranteeing that these controllers implemented with finite-precision arithmetic are safe. The finite precision arithmetic introduces the following errors, shown in Figure~\ref{fig:digitalsystem}. These errors are present in the real world, and we must model them accurately 
in order to capture the full behaviour of the system. Details of how we model this arithmetic are in Section~\ref{sec:}:

@@

\item The DAC could potentially introduce further errors. However, we assume that the DAC converts numbers to analogue signals with the same precision as the controller. 
As a result we can ignore any additional errors introduced by the DAC. Again, this assumption is reasonable based on commonly available hardware. It would, however, be straight forward to account for a DAC or ADC of different precision to the controller in our algorithm if necessary.
\end{itemize}

@@

We consider an input specification of the form 
$$
\exists P .\, \forall a.\, \phi(a, P),
$$  
where $P$ ranges over functions (where a function is represented by the program computing it),
$a$ ranges over ground terms, 
and $\phi$ is a quantifier-free logical formula. 
We~interpret the ground terms over some domain~$\mathcal{D}$. 
This is a synthesis problem, where the objective is to synthesise a valid $P$ that satisfies $\phi$ over all the $a$ terms.

@@

In the context of the formal synthesis of safe controllers, 
the set of possible {\sc inputs} corresponds to the set of possible initial states and the 
candidate program $P$ is a candidate controller $K$. The synthesis block generates a candidate
controller that works for a subset of the possible initial states, and the verifier checks whether 
the controller works for all possible initial states.

@@

%-------------------------------
\subsection{Jury's stability criteria} 
\label{ssec:stabspecification}
%-------------------------------

There are a number of procedures that can be used for stability analysis of dynamical models\note{~\cite{daes20161, Bessa16} is self-citation really the best citation for this statement??}.
%which can fit our automatic verification engine.  
%one based on Schur's decomposition
%and another one based on Jury's criterion~\cite{astrom1997computer}.  
Here we choose Jury stability criterion~\cite{astrom1997computer}, 
in view of its efficiency and ease of integration within our implementation.  
This method checks the stability working in the complex domain of the characteristic polynomial $S(z)$,  
%
% of the matrix $\left( \begin{array}{c} Y \\ U \end{array}\right)$, where $Y(z)$ is the closed-loop system output and U(z) is the controller output. 
%% \red{[unclear: what matrix? We need to clarify the relationship
%% between the char. poly. $S(z)$ and the later quantities $\Delta
%% N_{G}(z), \Delta D_{G}(z)$ and corresponding controller's
%% structure.]}.
%
considered in its general form as 
% for $S(z)$:
%
\begin{equation*}
S(z) = a_0z^N+a_1z^{N-1}+\cdots+a_{N-1}z+a_N,\,\, a_0\neq0. 
\end{equation*}
An LTI system is stable if all the roots of its characteristic polynomial $S(z)$ (i.e., the 
eigen values of the closed loop matrix $A_d-B_dK$) are inside the unit circle of the complex plane, 
i.e., their absolute values are less than one.

@@

\ifx\safetyrep
%-------------------------------
\subsection{Formal specification of safety} 
\label{ssec:safespecification}
%-------------------------------

@@

Furthermore, it is practically relevant to consider constraints $\phi_\mathit{input}$ on the input
signal $u_{k}$ and requirements $\phi_\mathit{init}$ on the initial states $x_0$, 
both of which we be expressed via given bounds as: 
$\phi_\mathit{input} := {\forall k\ge 0, \underline{u} \leq u_{k} \leq \overline{u}} $, 
and $\phi_\mathit{init} := \bigwedge_{i=1}^{n} \underline{x_{i,0}} \leq x_{i,0} \leq \overline{x_{i,0}}$. 
For the input signal, these requirements practically mean that the control input saturates in view of physical constraints. 
\fi

%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section{Numerical representation and soundness} 
\label{sec:numeric_rep}
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++

In this section we will quantify how the finite precision arithmetic in a digital controller affects the
safety and stability properties of an LTI system

\subsubsection{Safety of closed-loop models with finite precision controller error}
Let us first consider the effect of the quantization errors on safety. 
Within the controller, state values are manipulated at low precision,
alongside the vector multiplication $Kx$.
The inputs are computed using the following equation: 
%
\begin{align*}
u_{k}&=-(\mathcal{F}_{\langle I_c,F_c \rangle}(K)\cdot\mathcal{F}_{\langle I_c,F_c \rangle}(x_{k})). 
\end{align*}

This induces two types of the errors detailed above: first, the truncation
error due to representing $x_k$ as $\mathcal{F}_{\langle I_c,F_c
\rangle}(x_{k})$; and second, the rounding error introduced by the
multiplication operation. This error is behaviour modelled by the bounded model checking.

\subsubsection{Stability of closed-loop models with fixed-point controller error}
%-------------------------------
\label{sec:stab_FWL}
The proof of Jury's criterion~\cite{fadali} relies on the fact that the relationship between states and
next states is defined by $x_{k+1} = (A_d - B_dK) x_k$, all computed
at infinite precision.  When we employ a FWL digital controller, the
operation becomes:
%
\begin{align*}
x_{k+1} &= A_d \cdot x_{k} -(\mathcal{F}_{\langle I_c,F_c \rangle}(K)\cdot\mathcal{F}_{\langle I_c,F_c \rangle}(x_{k})).  \\
x_{k+1} &= (A_d  - B_dK) \cdot x_k + B_dK\delta, 
%\\ x_{k+1}&=A_d x_{k} + B_du_{k}
\end{align*}
%
where $\delta$ is the maximum error that can be introduced by the FWL
controller in one step, i.e., by reading the states values once and
multiplying by $K$ once.  We derive the closed form expression for $x_n$ as
follows:
%
\small
\begin{align*}
x_{1} &= (A_d  - B_dK)x_0 + B_dK\delta \\
x_{2} %&= (A_d  - B_dK)((A_d  - B_dK)x_0 + B_dK\delta ) + B_dK\delta \\
 &=(A_d  - B_dK)^2x_0 + (A_d  - B_dK)B_dK\delta + B_dK\delta \\
%x_{3}  &=(A_d  - B_dK)^3x_0 + (A_d  - B_dK)^2B_dK\delta + (A_d  - B_dK)B_dK\delta + B_dK\delta\\
x_{n} &= (A_d  - B_dK)^nx_0 + (A_d  - B_dK)^{n-1}B_dK\delta +\\
&\phantom{{}=1} ... + (A_d  - B_dK)^1B_dK \delta + B_dK\delta \\
  &= (A_d - B_dK)^nx_0 + \sum_{i=0}^{i=n-1}(A_d - B_dK)^iB_dk\delta. 
\end{align*}
\normalsize
The definition of asymptotic stability is that the system converges to a
reference signal, in this case we use no reference signal so an
asymptotically stable system will converge to the origin.  We know that the
original system with an infinite-precision controller is stable, because we
have synthesized it to meet Jury's criterion.  Hence, $(A_d - B_dK)^n x_0$ 
must converge to zero.

%T%o show that a fixed-point implementation of the same controller will
%converge to a finite value, we need to show that the infinite sum converges. 
The power series of matrices
converges~\cite{horn1990matrix} iff the eigenvalues of the matrix are less
than~1 as follows:
%
%\begin{align*}
$\sum_{i=0}^{\infty}T^i  = (I - T)^{-1}$, 
%\end{align*}
%
where $I$ is the identity matrix and $T$ is a square matrix. Thus, our system will converge to the value 
%
\begin{align*}
0 + (I - A_d + B_dK)^{-1}B_dk\delta \,. 
\end{align*}
%
As a result, if the value $(I - A_d + B_dK)^{-1}B_dk\delta$ is within the
safe space, then the synthesized fixed-point controller results in a safe
closed-loop model.  The convergence to a finite value, however, will not
make it asymptomatically stable.
However,
since in this paper we require stability only as a precursor to safety, it
is sufficient to check that the (perturbed, noisy) model converges to a
neighborhood of the equilibrium within the safe set.

We can thus disregard these steady-state errors (caused by
FWL effects) when stability is ensured by synthesis, and then verify its
safety accounting for the finite-precision errors.

@@

@@
